Determining the direction of causality in the face of measurement error

### 2016-02-14 docs/manuscript.Rmd
updated

With our ability to characterize the human phenome ever improving it is becoming increasingly common to use statistical tests to infer the causal relationships between correlated variables. A simple method for inferring if an exposure is on the causal pathway to an outcome is through mediation analysis, where the effect of an instrument on the outcome is tested before and after adjusting for the exposure. We show that in the face of measurement error this can lead to erroneous results, and that increasing sample size, rather than reducing bias, can often have the alarming effect of increasing certainty in the wrong answer. We argue that because measurement error is ubiquitous in phenotypic data, that mediation-based causal inference should be treated with caution. Finally, we demonstrate that Mendelian randomization is a method for causal inference that is robust to measurement error.

@@

### CIT direction

@@

Supposing there are two correlated variables, $x$ and $y$, with a variant $g$ associated with both. Assuming no horizontal pleiotropy (i.e. $g$ only has an effect on one of the variables through the other variable) it is desirable to know which of the variables it has a direct influence on. This can be achieved by testing for difference in the correlations $\rho_{g, x}$ and $\rho_{g, y}$ using Steiger's Z-test for correlated correlations within a population. It is calculated as

@@

@@

Here we consider a simple causal model where an exposure, $x$, has an effect on an outcome of interest, $y$, and that there is a known SNP, $g$, which has a direct effect on the exposure. Measurement error of an exposure can be modeled as some transformation of the true value that leads to the observed value, $x_O=f(x)$. For example, we can define $f(x) = \alpha_m + \beta_m x + \epsilon_m$, where $\alpha_m$ and $\beta_m$ influence the bias in the measurement of $x$, and $\epsilon_m$ represents the imprecision in the measurement of $x$. Here the true value of the exposure is partially explained by the genetic instrument, $g$, such that

@@

We performed simulations to verify that this problem does arise using the CIT method. `r cfig("cit_measurement_error_figure")` shows that when there is no measurement error in the exposure ($\rho_{x, x_O}=1$) the CIT is reliable in identifying the correct causal direction. However, as measurement error increases, eventually the CIT is more likely to infer a robust causal association in the wrong direction. Also of concern here is that increasing sample size does not solve the issue, indeed it only strengthens the evidence of the incorrect inference.

@@

If one is to select an instrument $g$ that has a direct influence on $x$ and $x$ is causally related to $y$, one can reason that the influence of $g$ on $y$ is the proportion of the effect of $x$ on $y$ that is explained by the effect of $g$ on $x$. Hence the effect estimate $\beta_{MR}$ of $x$ on $y$ is estimated as

$$
\beta_{MR} = \frac{\beta_y}{\beta_g} = \frac{cov(y, g)}{cov(x_O,g)} = \frac{cov(\beta_x x + \epsilon_x, g)}{cov(x + \epsilon_x, g)}
$$

@@

However, if we do not know whether the SNP $g$ has a direct influence on $x$ or $y$ then further efforts are required. If $x$ causes $y$ and $g$ is a valid instrument for $x$ then we expect that the Steiger test will obtain the correct result if the value $d = \rho_{g, x_O} - \rho_{g, y_O}$ is greater than 0. It can be shown that in the presence of measurement error, $d = \rho_{x, x_O} - \rho_{x,y}\rho_{y,y_O}$ (Appendix 2). Therefore, using the MR approach there are two potential advantages over the CIT. First, we can conduct a formal test for the direction of causality. Second the correct direction of causality can be inferred when measurement error in the exposure variable is present, as long as it is lower than the product of the measurement error in the outcome and the causal correlation between the exposure and the outcome. `r cfig("d_relationship_figure")` shows that in most cases, especially when the causal effect between $x$ and $y$ is not very high, the condition of $d > 0$ is satisfied.

We performed simulations to explore the performance of this approach in comparison to CIT. `r cfig("cit_mr_comparison_figure")` shows that, as predicted, when $d < 0$ the MR analysis is liable to infer the wrong direction of causality, and that this erroneous result is more likely to occur with increasing sample size. However, in all cases the MR analysis infers the wrong direction of causality at the same or a lower rate than the CIT. When $d > 0$ is satisfied we observe that in most cases the MR method has greater power to obtain evidence for causality than CIT, and always obtains the correct direction of causality.

@@

>>>>

With our ability to characterize the human phenome ever improving it is becoming increasingly common to use statistical tests to infer the causal relationships between correlated variables. A simple method for inferring if an exposure is on the causal pathway to an outcome is through mediation analysis, where the effect of an instrument on the outcome is tested before and after adjusting for the exposure. We show that in the face of measurement error this can lead to erroneous results, and that increasing sample size, rather than reducing bias, can often have the alarming effect of increasing certainty in the wrong answer. Commonly used tools that are predicated on this methodology such as the causal inference test (CIT) are shown to be susceptible. We demonstrate that Mendelian randomization (MR) is a method for causal inference that is robust to measurement error, and that a simple extension to MR can provide a method to estimate the direction of causality between correlated variables even when the biology of the instrument is unknown. We argue that because measurement error is ubiquitous in phenotypic data, mediation-based causal inference should be treated with caution.

@@

The analysis proceeds by simulating variables that have known causal effects and valid instruments, and then applying measurement error in different configurations. We then assess the performance of the CIT test and MR analysis for a) inference of a causal relationship and b) inferring the correct direction of causality. Included in the MR analysis is a simple extension that is used to make inference of the causal direction when it is unknown which of the correlated variables the instrument is directly acting upon. All analysis was performed using the R programming language [@RCoreTeam2015] and code is made available at [github url to go here]().

### CIT test

The CIT method [@Millstein2009] is implemented in the R package *R/cit* [@Millstein2016]. The *cit.cp* function was used to obtain an omnibus p-value of causality between an exposure $x$ and an outcome $y$ using a mediator $g$. To infer the direction of causality using the CIT method, the omnibus p-value was estimated for $x$ causing $y$, and for $y$ causing $x$, and the model that returned the lowest p-value was taken as the one denoting the correct causal direction.

@@

Supposing there are two correlated variables, $x$ and $y$, with a variant $g$ associated with both. Assuming no horizontal pleiotropy (i.e. $g$ only has an effect on one of the variables through the other variable) it is desirable to know which of the variables it has a direct influence on. This can be achieved by testing for a difference in the correlations $\rho_{g, x}$ and $\rho_{g, y}$ using Steiger's Z-test for correlated correlations within a population [@Steiger1980]. It is calculated as

@@

The $Z$ value is interpreted such that

$$
Z \left\{
\begin{array}{ll}
> 0, & x \to y\\
< 0, & y \to x\\
= 0, & x \perp\!\!\!\perp y 
\end{array} \right.
$$

and a p-value is generated from the $Z$ value to indicate confidence of difference in correlations $\rho_{gx}$ and $\rho_{gy}$. The test for a causal association is then obtained by performing two sample MR in the direction that is predicted to be the correct model based on the Z score test. For simplicity, the test is deemed to have obtained a robust association if the Z score test has a p-value $<  0.05$ and the two sample MR analysis has a p-value $< 0.05$. Note that the same approach can be applied to a two-sample MR setting, where

Simulations were conducting by creating variables of sample size $n$ for the exposure $x$, the measured values of the exposure $x_O$, the outcome $y$, the measured values of the outcome $y_O$ and the instrument $g$. In all models $x$ causes $y$ and $g$ is an instrument for $x$. Each variable was simulated such that:

$$
\begin{aligned}
g & \sim Binom(2, 0.5) \\
x & = \beta_g g + \epsilon_g \\
x_O & = \beta_{mx} x + \epsilon_{mx} \\
y & = \beta_x x + \epsilon_x \\
y_O & = \beta_{my} y + \epsilon_{my} \\
\end{aligned}
$$

All $\beta$ values were set to 1. Values of $\epsilon_*$ were generated such that

$$
\begin{aligned}
cor(g, x)^2 & = 0.1 \\
cor(x, y)^2 & = \{0.2, 0.4, 0.6, 0.8\} \\
var(\epsilon_{mx}) & = \{0, 0.2, 0.4, 0.6, 0.8, 1\} \\
var(\epsilon_{my}) & = \{0, 0.2, 0.4, 0.6, 0.8, 1\} \\
n & = \{100, 1000, 10000\}
\end{aligned}
$$

giving a total of 432 combinations of parameters. Each of these sets of variables was performed 100 times, and the CIT and MR methods were applied to each in order to evaluate the causal association of the simualted variables. Similar patterns of results were obtained for different values of $cor(g, x)$.

@@

Measurement error of an exposure can be modeled as some transformation of the true value that leads to the observed value, $x_O=f(x)$. For example, we can define $f(x) = \alpha_m + \beta_m x + \epsilon_m$, where $\alpha_m$ and $\beta_m$ influence the bias in the measurement of $x$, and $\epsilon_m$ represents the imprecision in the measurement of $x$. Here the true value of the exposure is partially explained by the genetic instrument, $g$, such that

@@

We performed simulations to verify that this problem does arise using the CIT method. `r cfig("cit_measurement_error_figure")` shows that when there is no measurement error in the exposure or outcome variables ($\rho_{x, x_O}=1$) the CIT is reliable in identifying the correct causal direction. However, as measurement error increases in the exposure variable, eventually the CIT is more likely to infer a robust causal association in the wrong direction. Also of concern here is that increasing sample size does not solve the issue, indeed it only strengthens the evidence of the incorrect inference.

@@

When selecting an instrument $g$ that has a direct influence on $x$ and $x$ is causally related to $y$, one can reason that the influence of $g$ on $y$ is the proportion of the effect of $x$ on $y$ that is explained by the effect of $g$ on $x$. Hence the effect estimate $\beta_{MR}$ of $x$ on $y$ is estimated as

$$
\beta_{MR} = \frac{\beta_y}{\beta_g} = \frac{cov(y, g)}{cov(x_O, g)} = \frac{cov(\beta_x x + \epsilon_x, g)}{cov(x + \epsilon_x, g)}
$$

@@

However, if we do not know whether the SNP $g$ has a direct influence on $x$ or $y$ then further efforts are required. If $x$ causes $y$ and $g$ is a valid instrument for $x$ then we expect that the Steiger test will obtain the correct result because the value $d = \rho_{g, x_O} - \rho_{g, y_O}$ is greater than 0 when the causal association between $x$ and $y$ is $\rho_{x, y} < 1$. But it can be shown that in the presence of measurement error, $d = \rho_{x, x_O} - \rho_{x,y}\rho_{y,y_O}$ (Appendix 2), thus it is important to note that under certain conditions when measurement error is present the Steiger test could make erroneous inference about causal direction because the value of $d$ is not restricted to being greater than 0. However, using the MR approach there are two potential advantages over the CIT. First, we can conduct a formal test for the direction of causality. Second the correct direction of causality can be inferred when measurement error in the exposure variable is present, as long as it is lower than the product of the measurement error in the outcome and the causal correlation between the exposure and the outcome. `r cfig("d_relationship_figure")` shows that in most cases, especially when the causal effect between $x$ and $y$ is not very large, the condition of $d > 0$ is satisfied.

We performed simulations to explore the performance of this approach in comparison to CIT. `r cfig("cit_mr_comparison_figure")` shows that, as predicted, when $d < 0$ the MR analysis is liable to infer the wrong direction of causality, and that this erroneous result is more likely to occur with increasing sample size. However, in all cases the MR analysis infers the wrong direction of causality at the same or a lower rate than the CIT. When $d > 0$ is satisfied we observe that in many cases the MR method has greater power to obtain evidence for causality than CIT, and always obtains the correct direction of causality.

@@

Researchers are often confronted with the problem of making causal inferences using a statistical framework on observational data, and unfortunately there are no solutions that work perfectly in all scenarios. In the face of measurement error it is evident that causal inference drawn through mediation analysis will be difficult to interpret, and this is a serious problem considering that it is often impossible to estimate the extent of measurement error for most experimental designs.

In the epidemiological literature this is relatively well understood and our initial analysis simply confirms that for widely used methods such as CIT they are indeed liable to the same issues as standard mediation based analysis. Specifically, we show that as measurement error in the exposure variable increases, CIT is likely to infer the wrong direction of causality. We also demonstrate that, though seemingly unintuitive, increasing sample size does not resolve the issue, rather it increases the significance of the incorrect inference.

Under many circumstances a practical solution to this problem is to use Mendelian randomisation instead of mediation based methods. Mendelian randomisation is robust in the face of measurement error and, given that the researcher has knowledge about the biology of the instrument being used in the analysis, can offer a direct solution to the issues that CIT faces. This assumption is often reasonable, for example SNPs are commonly used as instruments when they are found in genes with known biological relevance to the trait of interest. But on many occasions this is not the case, and it may be tempting to resort to using methods based on mediation in order to be able to both ascertain if there is a causal association and to infer the direction of causality. Here we have described a simple extension to MR which can be used as an alternative to mediation based methods. We show that this method is still liable to measurement error, but because it has different properties to the CIT it offers two main advantages. First, it uses a formal statistical framework to test for the robostness of the assumed direction of causality. Second, after testing in a comprehensive range of scenarios the MR based approach is much less likely to infer the wrong direction of causality compared to CIT.

There are a number of limitations to the analysis that we performed. We focused on the simple case of a single instrument in a single sample setting, and assumed that pleiotropy (the influence of the instrument on the outcome through a mechanism other than the exposure) was not present.

The overarching result from our simulations is that, regardless of the method used, inferring causal direction using an instrument of unknown biology is highly liable to measurement error.

Limitations

Pleiotropy

### 2016-02-14 docs/manuscript.tex
updated

\subsubsection{CIT direction}\label{cit-direction}

@@

Supposing there are two correlated variables, \(x\) and \(y\), with a
variant \(g\) associated with both. Assuming no horizontal pleiotropy
(i.e. \(g\) only has an effect on one of the variables through the other
variable) it is desirable to know which of the variables it has a direct
influence on. This can be achieved by testing for difference in the
correlations \(\rho_{g, x}\) and \(\rho_{g, y}\) using Steiger's Z-test
for correlated correlations within a population. It is calculated as

@@

@@

Here we consider a simple causal model where an exposure, \(x\), has an
effect on an outcome of interest, \(y\), and that there is a known SNP,
\(g\), which has a direct effect on the exposure. Measurement error of
an exposure can be modeled as some transformation of the true value that
leads to the observed value, \(x_O=f(x)\). For example, we can define
\(f(x) = \alpha_m + \beta_m x + \epsilon_m\), where \(\alpha_m\) and
\(\beta_m\) influence the bias in the measurement of \(x\), and
\(\epsilon_m\) represents the imprecision in the measurement of \(x\).
Here the true value of the exposure is partially explained by the
genetic instrument, \(g\), such that

@@

We performed simulations to verify that this problem does arise using
the CIT method. Figure 1 shows that when there is no measurement error
in the exposure (\(\rho_{x, x_O}=1\)) the CIT is reliable in identifying
the correct causal direction. However, as measurement error increases,
eventually the CIT is more likely to infer a robust causal association
in the wrong direction. Also of concern here is that increasing sample
size does not solve the issue, indeed it only strengthens the evidence
of the incorrect inference.

@@

If one is to select an instrument \(g\) that has a direct influence on
\(x\) and \(x\) is causally related to \(y\), one can reason that the
influence of \(g\) on \(y\) is the proportion of the effect of \(x\) on
\(y\) that is explained by the effect of \(g\) on \(x\). Hence the
effect estimate \(\beta_{MR}\) of \(x\) on \(y\) is estimated as

\[
\beta_{MR} = \frac{\beta_y}{\beta_g} = \frac{cov(y, g)}{cov(x_O,g)} = \frac{cov(\beta_x x + \epsilon_x, g)}{cov(x + \epsilon_x, g)}
\]

@@

However, if we do not know whether the SNP \(g\) has a direct influence
on \(x\) or \(y\) then further efforts are required. If \(x\) causes
\(y\) and \(g\) is a valid instrument for \(x\) then we expect that the
Steiger test will obtain the correct result if the value
\(d = \rho_{g, x_O} - \rho_{g, y_O}\) is greater than 0. It can be shown
that in the presence of measurement error,
\(d = \rho_{x, x_O} - \rho_{x,y}\rho_{y,y_O}\) (Appendix 2). Therefore,
using the MR approach there are two potential advantages over the CIT.
First, we can conduct a formal test for the direction of causality.
Second the correct direction of causality can be inferred when
measurement error in the exposure variable is present, as long as it is
lower than the product of the measurement error in the outcome and the
causal correlation between the exposure and the outcome. Figure 2 shows
that in most cases, especially when the causal effect between \(x\) and
\(y\) is not very high, the condition of \(d > 0\) is satisfied.

We performed simulations to explore the performance of this approach in
comparison to CIT. Figure 3 shows that, as predicted, when \(d < 0\) the
MR analysis is liable to infer the wrong direction of causality, and
that this erroneous result is more likely to occur with increasing
sample size. However, in all cases the MR analysis infers the wrong
direction of causality at the same or a lower rate than the CIT. When
\(d > 0\) is satisfied we observe that in most cases the MR method has
greater power to obtain evidence for causality than CIT, and always
obtains the correct direction of causality.

@@

@@

>>>>

\subsubsection{CIT test}\label{cit-test}

The CIT method\textsuperscript{1} is implemented in the R package
\emph{R/cit}\textsuperscript{22}. The \emph{cit.cp} function was used to
obtain an omnibus p-value of causality between an exposure \(x\) and an
outcome \(y\) using a mediator \(g\). To infer the direction of
causality using the CIT method, the omnibus p-value was estimated for
\(x\) causing \(y\), and for \(y\) causing \(x\), and the model that
returned the lowest p-value was taken as the one denoting the correct
causal direction.

@@

Supposing there are two correlated variables, \(x\) and \(y\), with a
variant \(g\) associated with both. Assuming no horizontal pleiotropy
(i.e. \(g\) only has an effect on one of the variables through the other
variable) it is desirable to know which of the variables it has a direct
influence on. This can be achieved by testing for a difference in the
correlations \(\rho_{g, x}\) and \(\rho_{g, y}\) using Steiger's Z-test
for correlated correlations within a
population\textsuperscript{\textbf{???}}. It is calculated as

@@

The \(Z\) value is interpreted such that

\[
Z \left\{
\begin{array}{ll}
> 0, & x \to y\\
< 0, & y \to x\\
= 0, & x \perp\!\!\!\perp y 
\end{array} \right.
\]

and a p-value is generated from the \(Z\) value to indicate confidence
of difference in correlations \(\rho_{gx}\) and \(\rho_{gy}\). The test
for a causal association is then obtained by performing two sample MR in
the direction that is predicted to be the correct model based on the Z
score test. For simplicity, the test is deemed to have obtained a robust
association if the Z score test has a p-value \(< 0.05\) and the two
sample MR analysis has a p-value \(< 0.05\). Note that the same approach
can be applied to a two-sample MR setting, where

Simulations were conducting by creating variables of sample size \(n\)
for the exposure \(x\), the measured values of the exposure \(x_O\), the
outcome \(y\), the measured values of the outcome \(y_O\) and the
instrument \(g\). In all models \(x\) causes \(y\) and \(g\) is an
instrument for \(x\). Each variable was simulated such that:

\[
\begin{aligned}
g & \sim Binom(2, 0.5) \\
x & = \beta_g g + \epsilon_g \\
x_O & = \beta_{mx} x + \epsilon_{mx} \\
y & = \beta_x x + \epsilon_x \\
y_O & = \beta_{my} y + \epsilon_{my} \\
\end{aligned}
\]

All \(\beta\) values were set to 1. Values of \(\epsilon_*\) were
generated such that

\[
\begin{aligned}
cor(g, x) & = \{0.01, 0.05, 0.1\} \\
cor(x, y) & = \{0.2, 0.4, 0.6, 0.8\} \\
var(\epsilon_{mx}) & = \{0, 0.2, 0.4, 0.6, 0.8, 1\} \\
var(\epsilon_{my}) & = \{0, 0.2, 0.4, 0.6, 0.8, 1\} \\
n & = \{100, 1000, 10000\}
\end{aligned}
\]

giving a total of 432 combinations of parameters. Each of these sets of
variables was performed 100 times, and the CIT and MR methods were
applied to each in order to evaluate the causal association of the
simualted variables.

@@

Measurement error of an exposure can be modeled as some transformation
of the true value that leads to the observed value, \(x_O=f(x)\). For
example, we can define \(f(x) = \alpha_m + \beta_m x + \epsilon_m\),
where \(\alpha_m\) and \(\beta_m\) influence the bias in the measurement
of \(x\), and \(\epsilon_m\) represents the imprecision in the
measurement of \(x\). Here the true value of the exposure is partially
explained by the genetic instrument, \(g\), such that

@@

We performed simulations to verify that this problem does arise using
the CIT method. Figure 1 shows that when there is no measurement error
in the exposure or outcome variables (\(\rho_{x, x_O}=1\)) the CIT is
reliable in identifying the correct causal direction. However, as
measurement error increases in the exposure variable, eventually the CIT
is more likely to infer a robust causal association in the wrong
direction. Also of concern here is that increasing sample size does not
solve the issue, indeed it only strengthens the evidence of the
incorrect inference.

@@

When selecting an instrument \(g\) that has a direct influence on \(x\)
and \(x\) is causally related to \(y\), one can reason that the
influence of \(g\) on \(y\) is the proportion of the effect of \(x\) on
\(y\) that is explained by the effect of \(g\) on \(x\). Hence the
effect estimate \(\beta_{MR}\) of \(x\) on \(y\) is estimated as

\[
\beta_{MR} = \frac{\beta_y}{\beta_g} = \frac{cov(y, g)}{cov(x_O, g)} = \frac{cov(\beta_x x + \epsilon_x, g)}{cov(x + \epsilon_x, g)}
\]

@@

However, if we do not know whether the SNP \(g\) has a direct influence
on \(x\) or \(y\) then further efforts are required. If \(x\) causes
\(y\) and \(g\) is a valid instrument for \(x\) then we expect that the
Steiger test will obtain the correct result because the value
\(d = \rho_{g, x_O} - \rho_{g, y_O}\) is greater than 0 when the causal
association between \(x\) and \(y\) is \(\rho_{x, y} < 1\). But it can
be shown that in the presence of measurement error,
\(d = \rho_{x, x_O} - \rho_{x,y}\rho_{y,y_O}\) (Appendix 2), thus it is
important to note that under certain conditions when measurement error
is present the Steiger test could make erroneous inference about causal
direction because the value of \(d\) is not restricted to being greater
than 0. However, using the MR approach there are two potential
advantages over the CIT. First, we can conduct a formal test for the
direction of causality. Second the correct direction of causality can be
inferred when measurement error in the exposure variable is present, as
long as it is lower than the product of the measurement error in the
outcome and the causal correlation between the exposure and the outcome.
Figure 2 shows that in most cases, especially when the causal effect
between \(x\) and \(y\) is not very large, the condition of \(d > 0\) is
satisfied.

We performed simulations to explore the performance of this approach in
comparison to CIT. Figure 3 shows that, as predicted, when \(d < 0\) the
MR analysis is liable to infer the wrong direction of causality, and
that this erroneous result is more likely to occur with increasing
sample size. However, in all cases the MR analysis infers the wrong
direction of causality at the same or a lower rate than the CIT. When
\(d > 0\) is satisfied we observe that in many cases the MR method has
greater power to obtain evidence for causality than CIT, and always
obtains the correct direction of causality.

@@

Researchers are often confronted with the problem of making causal
inferences using a statistical framework on observational data, and
unfortunately there are no solutions that work perfectly in all
scenarios. In the face of measurement error it is evident that causal
inference drawn through mediation analysis will be difficult to
interpret, and this is a serious problem considering that it is often
impossible to estimate the extent of measurement error for most
experimental designs.

In the epidemiological literature this is relatively well understood and
our initial analysis simply confirms that for widely used methods such
as CIT that it is indeed liable to the same issues as standard mediation
based analysis. Under many circumstances a practical solution to this
problem is to use Mendelian randomisation instead of mediation based
methods. Mendelian randomisation is robust in the face of measurement
error and, given that the researcher has knowledge about the biology of
the instrument being used in the analysis, can offer a direct solution
to the issues that CIT faces. This assumption is often reasonable, for
example exposures such as CRP

Pleiotropy

@@

22.Millstein, J. \emph{cit: Causal Inference Test}. (2016). at
\textless{}\url{http://cran.r-project.org/package=cit}\textgreater{}
